Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-01-23 chapters/chapter01.tex
implementing comments

%likelihood
As a statistical alternative to an optimality criterion, the likelihood had also been proposed for phylogenetic inference relatively early.
The idea of using maximum likelihood to estimate evolutionary trees was first proposed by \cite{Edwards1964}, although without a working implementation at the time of the publication.
Maximum likelihood is a probabilistic method of inference, that evaluates a hypothesis in terms of how probable it is in the light of the observed data.
In a phylogenetic setting, the likelihood function returns the probability that the proposed model and the proposed tree, understood as a topology and branch lengths, have generated the observed data.
The evolutionary history and associated set of evolutionary parameters with a higher probability is naturally preferred over a history with a lower probability.
By using optimization techniques one can find the parameter values that maximize the probability of observing the data.
In his influential work, \cite{Felsenstein1981} proposed an efficient algorithm for calculating a likelihood of sequence evolution on a tree under an evolutionary model of substitution, the so called `tree pruning' algorithm.
Unfortunately finding the maximum likelihood tree requires a search through the space of all possible tree topologies; in section \ref{sub:mlburden} we discuss in detail the computational burden associated with these calculations.
The effort required to exhaustively search through a space of all possible rooted trees grows explosively with the number of taxa, as depicted in Table \ref{tab:maxLikeBurden}.

@@

\begin{equation}
\left\{ P(t)\right\} _{ij}=\underset{k=0}{\overset{\infty}{\sum}}\mathbf{R}^{k}\frac{(r t)^{k}}{k!}e^{-r t}=\underset{k=0}{\overset{\infty}{\sum}}\frac{(\mathbf{R}r t)^{k}}{k!}e^{-r t},
\label{eq:matrixExp1}
\end{equation}

\noindent
where $\left\{ P(t)\right\} _{ij}$ denotes the $ij\text{-th}$ element of the transition probability matrix $\mathbf{P}$ over time $t \geq 0$.
Let us denote the matrix $\mathbf{R}-\mathbf{I}$, where $\mathbf{I}$ is a $K \times K$ identity matrix, by $\mathbf{Q}$.
Using the power series definition of the transition probability matrix: $\exp(\mathbf{R})=\underset{k=0}{\overset{\infty}{\sum}}\mathbf{R}^{k}/k!$, we can write Equation~(\ref{eq:matrixExp1}) as:

\begin{equation}
\left\{ P(t)\right\} _{ij}=e^{\mathbf{R}r t}e^{-r t} =e^{(\mathbf{R}-\mathbf{I})r t}=e^{\mathbf{Q}r t} 
\label{eq:matrixExp2}
\end{equation}

@@

\begin{align}
\mathbf{P}\left(t\right) &=& \left[\begin{array}{ccccc}
p_{0}(t) & p_{1}(t) & \cdots & \cdots & p_{1}(t)\\
p_{1}(t) & p_{0}(t) &  &  & \vdots\\
\vdots &  & \ddots & p_{0}(t) & p_{1}(t)\\
p_{1}(t) & \cdots & \cdots & p_{1}(t) & p_{0}(t)
\end{array}\right],  \nonumber \\
&\text{where}& \ensuremath{\begin{cases}
p_{0}(t)=\frac{1}{4}+\frac{3}{4}e^{-4rt}\\
p_{1}(t)=\frac{1}{4}-\frac{1}{4}e^{-4rt}
\end{cases}}.
\label{eq:jc69Finite}
\end{align}

@@

\noindent
where $\mathbf{\Theta}$ represents the tree topology, the branch lengths, the parameters of the substitution model and other parameters of interest that we want to estimate.
This quantity expresses how likely it is to observe the data given a specific value of the parameter(s) of interest $\mathbf{\Theta}$.
By finding the value of $\mathbf{\Theta}$ that maximizes (\ref{eq:likelihood}) we obtain an estimate of the most plausible hypothesis for the data.

@@

The prior choice is perhaps one of the most controversial points of the Bayesian inference.
The prior distributions may come from the preceding analysis of other sets of data, or they may form some set of biologically plausible values.
However in some cases there may be no suitable objective distribution at hand, in which case one often resorts to the so-called \emph{vague} or \emph{diffuse} priors, which have large variances.
All of these proper prior choices lead to a proper posterior distribution, yet some controversies about their choice may still arise.

@@

Figure~\ref{fig:bayes1} shows the resulting posterior distribution along with prior and scaled likelihood.
This illustrates that the data `pulls' the posterior away from the prior, and the strength by which is does so depends on its variance, resulting in a posterior centered around 0.06 in this case.
In this example, the posterior distribution can be obtained analytically because the beta distribution is a so-called \emph{conjugate prior} for the binomial model, resulting in the posterior that is also beta-distributed.

@@

The Metropolis-Hasting algorithm, as proposed by \citet{Metropolis1953}, is an MCMC algorithm for sampling from multi-dimensional distributions.
It proceeds by constructing a Markov chain with its states being the values of the parameter of interest $\theta$ and whose equilibrium distribution is the posterior $P\left(\theta|\mathbf{X}\right)$ as defined in Equation~\ref{eq:bayesian1}.
After choosing an arbitrary starting state for the chain, new candidate states $\theta^{*}$ are generated from the proposal density $q(\theta^{*} | \theta[t])$, also called the kernel.

@@

$$\theta^{*}\sim U\left[\theta[t]-\omega,\theta[t]+\omega\right],$$

@@

\begin{algorithm}[H]
\centering
\begin{algorithmic}[1]
% \footnotesize{
%
\State Pick a starting value of the chain $\theta \left[ t \right]$ for $t=0$.
%
\For{int $\left(t=0; \; t<\text{chain length}; \; t++\right)$}
%
\State Generate a candidate state from the proposal distribution (kernel):
$$\theta^{*} \sim q(\theta[t])$$
%
\State Calculate the acceptance ratio:
$$r=\frac{prior\left(\theta^{*}\right)}{prior\left(\theta\left[t\right]\right)}\cdot\frac{Likelihood\left(\mathbf{X}|\theta^{*}\right)}{Likelihood\left(\mathbf{X}|\theta\left[t\right]\right)}\cdot\frac{q\left(\theta\left[t\right]|\theta^{*}\right)}{q\left(\theta^{*}|\theta\left[t\right]\right)}.$$
%
\State Sample $u\sim U[0,1]$.
%
\If{$\left( u < r \right)$}
%
\State Set $\theta[t+1]=\theta^*.$
%
\Else 
%
\State Set $\theta[t+1]=\theta[t].$
%
\EndIf
%
\EndFor
% }
\end{algorithmic}
\caption{
{ \footnotesize 
{\bf The Metropolis-Hastings algorithm} 
}% END: footnotesize
}
\label{alg:metropolisHastings}
\end{algorithm}

>>>>

%likelihood
As a statistical alternative to an optimality criterion, the likelihood had also been proposed for phylogenetic inference relatively early.
The idea of using maximum likelihood to estimate evolutionary trees was first proposed by \cite{Edwards1964}, although without a working implementation at the time of the publication.
Maximum likelihood is a probabilistic method of inference, that evaluates a hypothesis in terms of how probable it is in the light of the observed data.
In a phylogenetic setting, the likelihood function returns the probability that the proposed model and the proposed tree, understood as a topology and branch lengths, have generated the observed data.
\myedit{svthree}{
In general an evolutionary history and associated set of evolutionary parameters with higher probability of generating the observed data is naturally preferred over a history with a lower probability of generating it.
By using optimization techniques one can find the parameter values that maximize the probability of observing the data.
}
% The evolutionary history and associated set of evolutionary parameters with a higher probability is naturally preferred over a history with a lower probability.
% By using optimization techniques one can find the parameter values that maximize the probability of observing the data.
In his influential work, \cite{Felsenstein1981} proposed an efficient algorithm for calculating a likelihood of sequence evolution on a tree under an evolutionary model of substitution, the so called `tree pruning' algorithm.
Unfortunately finding the maximum likelihood tree requires a search through the space of all possible tree topologies; in section \ref{sub:mlburden} we discuss in detail the computational burden associated with these calculations.
The effort required to exhaustively search through a space of all possible rooted trees grows explosively with the number of taxa, as depicted in Table \ref{tab:maxLikeBurden}.

@@

\begin{equation}
\left\{ P(t) \right\} _{ij}= \left\{ \underset{k=0}{\overset{\infty}{\sum}}\mathbf{R}^{k}\frac{(r t)^{k}}{k!}e^{-r t} \right\} _{ij} = \left\{ \underset{k=0}{\overset{\infty}{\sum}}\frac{(\mathbf{R}r t)^{k}}{k!}e^{-r t} \right\} _{ij},
\label{eq:matrixExp1}
\end{equation}

\noindent
where $\left\{  \cdot \right\} _{ij}$ denotes the $ij\text{-th}$ element of a matrix.
% the transition probability matrix $\mathbf{P}$ over time $t \geq 0$.
Let us denote the matrix $\mathbf{R}-\mathbf{I}$, where $\mathbf{I}$ is a $K \times K$ identity matrix, by $\mathbf{Q}$.
Using the power series definition of the transition probability matrix: $\exp(\mathbf{R})=\underset{k=0}{\overset{\infty}{\sum}}\mathbf{R}^{k}/k!$, we can write Equation~(\ref{eq:matrixExp1}) as:

\begin{equation}
\left\{ \left\{ P(t)\right\} _{ij} \right\} _{ij} = \left\{ e^{\mathbf{R}r t}e^{-r t} \right\} _{ij} = \left\{ e^{(\mathbf{R}-\mathbf{I})r t} \right\} _{ij}= \left\{ e^{\mathbf{Q}r t} \right\} _{ij}
\label{eq:matrixExp2}
\end{equation}

@@

\begin{align}
% \mathbf{P}\left(t\right) &=& \left[\begin{array}{cccc}
% p_{0}(t) & p_{1}(t) & p_{1}(t) & p_{1}(t)\\
% p_{1}(t) & p_{0}(t) & p_{1}(t) & p_{1}(t)\\
% p_{1}(t) & p_{1}(t) & p_{0}(t) & p_{1}(t)\\
% p_{1}(t) & p_{1}(t) & p_{1}(t) & p_{0}(t)
% \end{array}\right]
%  ,  \nonumber \\
% &\text{where}& \ensuremath{\begin{cases}
% p_{0}(t)=\frac{1}{4}+\frac{3}{4}e^{-4rt}\\
% p_{1}(t)=\frac{1}{4}-\frac{1}{4}e^{-4rt}
% \end{cases}}.
p_{ij}\left(t\right)=\begin{cases}
\frac{1}{4}+\frac{3}{4}e^{-4rt} & \text{if }i=j\\
\frac{1}{4}-\frac{1}{4}e^{-4rt} & \text{otherwise}
\end{cases}
\label{eq:jc69Finite}
\end{align}

@@

\noindent
where $\mathbf{\Theta}$ represents the tree topology, the branch lengths, the parameters of the substitution model and other parameters of interest that we want to estimate.
This quantity expresses how likely it is to observe the data given a specific value of the parameter(s) of interest $\mathbf{\Theta}$.
\myedit{sveight}{
By finding the value of $\mathbf{\Theta}$ that maximizes (\ref{eq:likelihood}) we obtain an estimate of the most plausible 
%hypothesis for the data.
set of parameters given observed data.}

@@

\myedit{svnine}{
The prior choice is perhaps one of the most controversial points of the Bayesian inference.
The prior distributions may come from the preceding analysis of other sets of data, or they may form some set of biologically plausible values.
However in some cases there may be no suitable objective distribution at hand, in which case one often resorts to the so-called \emph{vague} or \emph{diffuse} priors, which have large variances.
Some popular uninformative priors on continuous, unbounded variables may be improper, \textit{i.e.} their distribution is infinitesimal on an infinite range in order to integrate to $1$.
Among examples of such priors is the uniform distribution on an infinite interval.
Although many of of these prior choices sill lead to a proper posterior distribution, yet some controversies about their choice may still arise.
}

@@

Figure~\ref{fig:bayes1} shows the resulting posterior distribution along with prior and scaled likelihood.
The likelihood of the binomial model used in this example is simply:

\begin{align}
L\left(\theta|x\right)=\left(\begin{array}{c}
N\\
x
\end{array}\right)\theta^{x}(1-\theta)^{N-x}
\label{eq:binomial_like}
\end{align}

Equation~\ref{eq:binomial_like} is plotted in Figure~\ref{fig:bayes1} over a range of values of $\theta$, resulting in the black filled distribution.
The Beta distribution we chose for the prior has a following distribution function:

\begin{align}
p(\theta)=\frac{1}{B\left(\alpha,\beta\right)}\theta^{\alpha-1}(1-\alpha)^{\beta-1}
\label{eq:beta_dist}
\end{align}

Equation~\ref{eq:beta_dist} corresponds to the white filled curve in in Figure~\ref{fig:bayes1}.
Following the Bayes rule from Equation~\ref{eq:bayesian1} we can combine the likelihood and prior to form the posterior, which then becomes:

\begin{eqnarray}
p\left(\theta|x\right)=& \frac{\left(\begin{array}{c}
N\\
x
\end{array}\right)\theta^{x}(1-\theta)^{N-x}\times\frac{1}{B\left(\alpha,\beta\right)}\theta^{\alpha-1}\left(1-\theta\right)^{\beta-1}}{\int\left(\begin{array}{c}
N\\
x
\end{array}\right)\theta^{x}(1-\theta)^{N-x}\times\frac{1}{B\left(\alpha,\beta\right)}\theta^{\alpha-1}\left(1-\theta\right)^{\beta-1}d\theta} \nonumber \\
=& \frac{1}{B\left(\alpha+x,\beta+N-x\right)}\theta^{\alpha+x-1}(1-\theta)^{\beta+N-x-1}
\label{eq:beta_binom_post}
\end{eqnarray}

Equation~\ref{eq:beta_binom_post} appears as a grey filled area in Figure~\ref{fig:bayes1}.
This illustrates that the data `pulls' the posterior away from the prior, and the strength by which is does so depends on its variance, resulting in a posterior centered around 0.06 in this case.
In this example, the posterior distribution can be obtained analytically because the beta distribution is a so-called \emph{conjugate prior} for the binomial model, resulting in the posterior that is also beta-distributed.

@@

The Metropolis-Hasting algorithm, as proposed by \citet{Metropolis1953}, is an MCMC algorithm for sampling from multi-dimensional distributions.
It proceeds by constructing a Markov chain with its states being the values of the parameter of interest $\theta$ and whose equilibrium distribution is the posterior $P\left(\theta|\mathbf{X}\right)$ as defined in Equation~\ref{eq:bayesian1}.
After choosing an arbitrary starting state for the chain, new candidate states $\theta^{*}$ are generated from the proposal density, also called the kernel, $q(\theta^{*} | \theta[t]), \; t=1,\ldots,M$, where $M$ is the chain length and $\theta[t]$ is the current value of the chain.

@@

$$\theta^{*}\sim U\left[\theta[t]-\omega,\theta[t]+\omega\right], \; t=1,\ldots,M,$$

@@

\begin{algorithm}[H]
\centering
\begin{algorithmic}[1]
% \footnotesize{
%
\State Pick a starting value of the chain $\theta \left[ t \right]$ for $t=0$.
%
\For{int $\left(t=1; \; t<M-1; \; t++\right)$}
%
\State Generate a candidate state from the proposal distribution (kernel):
$$\theta^{*} \sim q(\theta[t])$$
%
\State Calculate the acceptance ratio:
$$r=\frac{prior\left(\theta^{*}\right)}{prior\left(\theta\left[t\right]\right)}\cdot\frac{Likelihood\left(\mathbf{X}|\theta^{*}\right)}{Likelihood\left(\mathbf{X}|\theta\left[t\right]\right)}\cdot\frac{q\left(\theta\left[t\right]|\theta^{*}\right)}{q\left(\theta^{*}|\theta\left[t\right]\right)}.$$
%
\State Sample $u\sim U[0,1]$.
%
\If{$\left( u < r \right)$}
%
\State Set $\theta[t+1]=\theta^*.$
%
\Else 
%
\State Set $\theta[t+1]=\theta[t].$
%
\EndIf
%
\EndFor
% }
\end{algorithmic}
\caption{
{ \footnotesize 
{\bf The Metropolis-Hastings algorithm} 
}% END: footnotesize
}
\label{alg:metropolisHastings}
\end{algorithm}

### 2015-01-23 frontback/acknowledgments.tex
addressed comment on the dated tree tips
implementing comments

\paragraph{} 
To all my past and present friends from the Evolutionary and Computational Virology laboratory, Bram Vrancken, Guy Baele, Nidia Trovao, Nuno Faria for all the good work, good times, good beers and jenevers we had together.

@@

My friends from Poland, who I don't see as often as I would like to, but when we meet it is always something to remember. 
Wojtkowi - za pogod\k{e} ducha. 
Adamowi - \.{z}eby\'{s} zawsze wszystkich przechytrzy\l{}.
Kamilowi - uwa\.{z}aj na p\l{}oty. 
Ma\'{c}kowi - za wsp\'{o}lne wspinanie.
Marcinowi - widzimy si\k{e} za rzadko, wi\k{e}c pijemy za szybko.

\emph{`In the moment when I'm on the summit there is no explosion of happiness - happiness is something experienced when everything remains in front of you, when you know that to reach the peak you have only a few hundred, a few dozen meters, when it is right in front of you. This is the time of happiness.'} 
---Jerzy Kukuczka

>>>>

\paragraph{} 
To all my friends from the Evolutionary and Computational Virology laboratory, Bram Vrancken, Guy Baele, Nidia Trovao, Joakim Esbj\"ornsson and Nuno Faria for all the good work, good times, good beers and jenevers we had together.
To all the people from the Rega Institute, for all the help and support I have received, especially to Prof. Anne-Mieke Vandamme for testing my presentation skills, Yoeri Schrooten for evenings we spent with PlayStation and caipirinhas, Ria Swinnen for all the good suggestions on where to go out in Leuven, but also to Kristof Theys, Pieter Libin and Sietse Huysmans.

@@

My friends from Poland, who I don't see as often as I would like to, but when we meet it is always something to remember.

Wojtkowi - za pogod\k{e} ducha. 
Kamilowi i Marcinowi - za zawsze udane wyj\'{s}cia na miasto.
Adamowi i Ma\'{c}kowi - za wsp\'{o}lne wspinanie.

Jerzy Kukuczka once said:
\emph{`In the moment when I'm on the summit there is no explosion of happiness - happiness is something experienced when everything remains in front of you, when you know that to reach the peak you have only a few hundred, a few dozen meters, when it is right in front of you. This is the time of happiness.'} 
%---Jerzy Kukuczka
